Lossy compression of DNA sequencing quality data

### 2015-05-13 manuscript/q-compression.tex
Added text edits.

Quality values can be binned, and the minimum number of bins that
allows for a any distinction among quality values is two; i.e., two
categories of quality: ``good'' and ``bad''. We implement 2-bin
encoding by setting a quality value threshold empirically determined
by the distribution of quality values across reads. Base-calls are
marked ``bad'' if their quality value falls below the first quartile
minus 1.5 $\times$ the interquartile range (IQR), which is the
difference between the first and third quartile. 1.5 $\times$ IQR is
the value used by Tukey's box plot~\citep{mcgill1978variations}. The
main benefit of this approach is that it is completely data-dependent,
and no assumptions regarding the distribution of the quality values
need to be made.

@@

The economic costs of memory use for compression with
\textsc{QualComp}, in in our interpretation, appear to have no fixed
costs, and the marginal costs that are the product of the number of
reads times the cost of representing the model parameters.

@@

As an additional point of reference regarding compression
effectiveness, the recently published
Read-Quality-Sparsifier~\citep{DBLP:conf/recomb/YuYB14} achieved
best-case compression of 0.254 bits/base-call, and a mean compression
of 1.841 bits/base-call; both values are higher than all of the
profile results and for some results of other methods presented here
(Figure~\ref{fig:mse_vs_bpbp}).

@@

Preprocessing involves two steps: \emph{discarding} reads that are
deemed to be poor-quality overall, and \emph{trimming} the
poor-quality regions of the reads. After trimming, the majority of
compression methods retain more base-calls than the uncompressed
sequences (Figure \ref{fig:preprocessing}). In general, as a given
compression method increases in complexity --- i.e., as the number of
profiles, polynomial degrees, or rate increases --- the amount of
base-calls retained more closely approximates the number of base-calls
retained using the uncompressed sequences. The compression methods on
the \textit{Mus musculus} data set had the greatest proportion of
retained base-calls compared to the uncompressed sequences. The
\textit{Escherichia coli} MiSeq data set had the smallest range.

>>>>

Quality values can be binned, and the minimum number of bins that
allows for any distinction among quality values is two; i.e., two
categories of quality: ``good'' and ``bad''. We implement 2-bin
encoding by setting a quality value threshold empirically determined
by the distribution of quality values across reads. Base-calls are
marked ``bad'' if their quality value falls below the first quartile
minus 1.5 $\times$ the interquartile range (IQR), which is the
difference between the first and third quartile. 1.5 $\times$ IQR is
the value used by Tukey's box plot~\citep{mcgill1978variations}. The
main benefit of this approach is that it is completely data-dependent,
and no assumptions regarding the distribution of the quality values
need to be made.

@@

The economic costs of memory use for compression with
\textsc{QualComp}, in our interpretation, appear to have optional fixed
costs for storing the cluster representatives, and the marginal costs that are the product of the number of
reads times the cost of representing the model parameters.

@@

As an additional point of reference regarding compression
effectiveness, the recently published
Read-Quality-Sparsifier~\citep{DBLP:conf/recomb/YuYB14} achieved
best-case compression of 0.254 bits/base-call on \textit{Homo sapiens} chromosome 21 data (read lengths ranging from 50-110 nt), and a mean compression
of 1.841 bits/base-call; both values are higher than all of the
profile results and for some results of other methods presented here
(Figure~\ref{fig:mse_vs_bpbp}).

@@

Preprocessing involves two steps: \emph{discarding} reads that are
deemed to be poor-quality overall, and \emph{trimming} the
poor-quality regions of the reads. After trimming, the majority of
compression methods retain more base-calls than the uncompressed
sequences (Figure \ref{fig:preprocessing}). In general, as a given
compression method increases in complexity --- i.e., as the number of
profiles, polynomial degrees, or rate increases --- the amount of
base-calls retained more closely approximates the number of base-calls
retained using the uncompressed sequences. The compression methods on
the \textit{Mus musculus} data set has the greatest proportion of
retained base-calls compared to the uncompressed sequences. The
\textit{Escherichia coli} MiSeq data set has the smallest range.
